Nonlinear Dendritic Coincidence Detection for Supervised Learning

### 2021-05-09 notes/frontiers_manuscript/frontiers.tex
.

\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\graphicspath{{./figures/}}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[left=1.5cm, right=1.5cm, top=2.50cm, bottom=2.50cm]{geometry}
\usepackage{multicol}
\usepackage{float}

\author{Fabian Schubert}
\title{Nonlinear Dendritic Coincidence Detection for Supervised Learning}

\usepackage[numbers]{natbib}
\bibliographystyle{nourlplainnat.bst}
%\bibliographystyle{abbrvnat}

\renewcommand{\arraystretch}{1.5}

\usepackage{layouts}

\begin{document}
	\maketitle

		\section{Introduction}

@@

		\section{Model}

		\subsection{Neuron Model}

		The neuron model used throughout this study is a discrete-time rate encoding model that
		uses two separate input variables, subsuming the total synaptic
		input current injected arriving at the basal (proximal) and apical (distal) 
		dendritic structure of a pyramidal neuron, respectively. The model is a slightly
		simplified version of a phenomenological model proposed by \citet{Shai_2015}.
		Denoting the input currents by $I_{\rm p}$ (proximal) and $I_{\rm d}$ (distal), 
		the model is written as
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{align}
		\begin{split}
		y\left(t\right) &= \alpha  \sigma\left( I_{\rm p}(t) - \theta_{p0} \right)
		\left[1-\sigma\left(I_{\rm d}(t) - \theta_d\right)\right] \\
		&+ \sigma\left(I_{\rm d}(t) - \theta_d \right)\sigma\left( I_{\rm p}(t) - \theta_{p1} \right)
		\end{split} \label{eq:comp_model}\\
		\sigma(x) &\equiv \frac{1}{1+\exp(-4x)} \; .
		\end{align}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		Here, $\theta_{p0}$, $\theta_{p1}$ and $\theta_d$ are threshold variables with
		respect to proximal and distal input. Overall, this equation describes two distinct
		regions of neural activation in the $(I_{\rm p}, I_{\rm d})$-space which differ in their
		maximal firing rates, which are set to $1$ and $\alpha$, where $0 < \alpha < 1$.
		A plot of \eqref{eq:comp_model} is shown in Fig.~\ref{fig:comp_model}.
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\centering
			\includegraphics[width=0.8\columnwidth]{plot_comp_mod_marks.png}
			\caption{{\bf Two-compartment rate model.} Firing rate as a
			function of proximal and distal input $I_{\rm p}$ and $I_{\rm d}$, 
			see \eqref{eq:comp_model}. 
			The thresholds $\theta_{p0}$, $\theta_{p1}$ and $\theta_d$ define
			two regions of neural activity with a a maximal firing rate of $1$
			and $\alpha$.}
			\label{fig:comp_model}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

		\subsection{Plasticity}

		We implemented a Hebbian plasticity rule for the basal 
		synaptic weights given by the following update equation:
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{align}
			w_i(t+1) &= w_i(t) + \mu_w \left(x_{p,i}(t) - \tilde{x}_{p,i}(t)\right)
			\left(y(t) - \tilde{y} \right)
			\label{eq:hebb_plast} \\
			\tilde{x}_{p,i}(t+1) &= (1-\mu_{av})\tilde{x}_{p,i}(t) + \mu_{av}x_{p,i}(t+1) \\
			\tilde{y}(t+1) &= (1-\mu_{av})\tilde{y}(t) + \mu_{av}y(t+1)
		\end{align}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		Additionally, we use a
		synaptic normalization constraint
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			w_i(t) \rightarrow \frac{w_i(t)}{\lVert \mathbf{w}(t)\rVert}
			\label{eq:weight_norm}
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		in each time step, where $\lVert \mathbf{w}(t)\rVert$ denotes the Euclidean
		norm of the synaptic weight vector. 
%		The threshold $\theta_y$ was
%		chosen such that LTP is present if neural activity is in the
%		high-activity regime (yellow area in Fig.~\ref{fig:comp_model}), while
%		LTD is present in the low-activity regime 
%		(given by $\alpha$, see Fig.~\ref{fig:comp_model}). Hence, we set
%		$\theta_y = (1+\alpha)/2$.

		For comparative reasons, the point neuron model is equipped with the
		same plasticity rule for the proximal weights as \eqref{eq:hebb_plast}. 
%		except that the threshold $\theta_y$ was dynamically adjusted as
%		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%		\begin{equation}
%			\theta_y(t+1) = \left(1-\mu_{av}\right)\theta_y(t) + \mu_{av} y^2(t) \; , 
%		\end{equation}
%		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%		which is an implementation of a running average of the square activity,
%		as used in the classic BCM-rule.

@@

		\section{Results}
		\label{sect:results}

		\subsection{Increased Alignment between Basal and Apical Inputs}
		\label{sect:alignment}

@@

		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\includegraphics[width=\columnwidth]{corr_dimension_scaling_high_input_dim}
			\caption{{\bf Alignment between Basal and Apical Input.} Color
			encodes the Pearson correlation $\rho[I_p,I_d]$ for different
			number of orthogonal distraction directions $N_{dist}$ 
			and the corresponding scaling faction $s$.}
			\label{fig:corr_dimension_scaling}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		\subsection{Supervised Learning in a Linear Classification Task}
		\label{sect:classification}

		Next, we investigated if the observed differences would also improve
		the performance in an actual supervised learning task.
		For this purpose, we constructed presynaptic basal input as illustrated
		in Fig.~\ref{fig:illustration_classification}.
		Written in vector form,
		each sample from the basal input is generated from the following expression:
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{equation}
			\mathbf{x}_p(t) = \mathbf{b} + \mathbf{a}\left(c(t) + \sigma_a \zeta_a(t) \right) 
			+ s \cdot \sum_{i=1}^{N_{dist}} \zeta_{dist,i}(t) \mathbf{v}_{dist,i} \; .
		\end{equation}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		Here, $\mathbf{b}$ is a random vector drawn uniformly from
		$(0,1)^N$, $\mathbf{a}$ is random unit vector as introduced in 
		Section~\ref{sect:alignment}, $c(t)$ is a binary variable drawn 
		from $\{-0.5,0.5\}$ with equal probability and $\zeta_a(t)$ and the
		$\zeta_{dist,i}(t)$ are independent random Gaussian variables with 
		zero mean and unit variance. 
		Hence, $\sigma_a$ simply denotes the standard deviation of each Gaussian
		cluster along the direction of the normal vector $\mathbf{a}$ and was
		set to $\sigma_a = 0.25$. 
		Finally, the set of $\mathbf{v}_{dist,i}$ forms a randomly generated
		orthogonal basis of $N_{dist}$ unit vectors which---just as in 
		Section~\ref{sect:alignment}---are also orthogonal to $\mathbf{a}$.
		The free parameter $s$ parameterized
		the standard deviation along this subspace orthogonal to $\mathbf{a}$.
		As indicated by the time dependence, the Gaussian and binary random
		variables are drawn for each time step. The vectors
		$\mathbf{b}$, $\mathbf{a}$, and $\mathbf{v}_{dist,i}$ are generated
		once before the beginning of a simulation run.
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\centering
			\includegraphics[width=0.75\columnwidth]{illustration_classification}
			\caption{{\bf Input Space for a Simple Linear Classification Task.}
				Two clusters of presynaptic basal activity were generated from 
				multivariate Gaussian distributions. Here, $s$ denotes the standard
				deviation orthogonal to the normal vector of the classification
				hyperplane.}
			\label{fig:illustration_classification}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		For the classification task, we set up two output neurons receiving
		the same basal presynaptic input, while the top-down input encoded
		the correct linear classification in a one-hot scheme, that is
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{align}
			x_{d,0}(t) &= 1 - \Theta\left( \left(\mathbf{x}_p(t) - \mathbf{b}\right)^T \mathbf{a}\right) \\
			x_{d,1}(t) &= \Theta\left( \left(\mathbf{x}_p(t) - \mathbf{b}\right)^T \mathbf{a}\right) \; ,
		\end{align}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		where $\Theta(x)$ is the Heaviside step function.

		As in the previous experiment, we ran a full simulation until 
		all dynamic variables reached a stationary state. After this,
		a test run without plasticity and with the apical input turned off 
		was used to evaluate the classification	performance. 
		For each sample,
		the index of the neuron with the highest activity was used
		as the predicted class. Accuracy was then calculated as the fraction
		of correctly classified samples.
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\includegraphics[width=\columnwidth]{classification_dimension_scaling_high_input_dim}
			\caption{{\bf Binary Classification Accuracy.}
			Fraction of correctly classified patterns as illustrated in
			Fig.~\ref{fig:illustration_classification}, see 
			Section~\ref{sect:classification}.¸}
		\label{fig:classification_accuracy}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		The resulting accuracy as a function of $N_{dist}$ and $s$
		is shown in Fig.~\ref{fig:classification_accuracy}.
		Albeit differences are small, the compartment model does
		show a better overall accuracy for the tested parameter range.
		It should be noted, though, that the advantage of the compartment
		model becomes even less prominent when looking at the actual
		correlation between proximal and distal input as a
		measure of successful learning (as done in the previous section).
		Fig.~\ref{fig:classification_correlation} in the supplementary 
		section shows that the compartment model allows for a marginally
		larger scaling among the distraction dimensions.

		\subsection{Non-Hebbian Learning Rules}
		\label{sect:non-hebbian}

@@

		The compartment model also significantly benefits from
		the BCM rule in terms of basal-apical alignment as tested in Sect.~\ref{sect:alignment}, 
		while only marginal
		improvements can be observed for the point model
		(compare Fig.~\ref{fig:corr_dimension_scaling_bcm} 
		with Fig.~\ref{fig:corr_dimension_scaling}).

		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\includegraphics[width=\columnwidth]{classification_dimension_scaling_bcm_high_input_dim}
			\caption{{\bf Binary Classification Accuracy, BCM Rule.}
				Fraction of correctly classified patterns as illustrated in
				Fig.~\ref{fig:illustration_classification}, after training with
				a BCM-like learning rule.¸}
			\label{fig:classification_accuracy_bcm}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\includegraphics[width=\columnwidth]{corr_dimension_scaling_bcm_high_input_dim}
			\caption{{\bf Alignment between Basal and Apical Input, BCM Rule.}
				Color encodes the Pearson correlation $\rho[I_p,I_d]$ for different
				number of orthogonal distraction directions $N_{dist}$ 
				and the corresponding scaling faction $s$ after training
				with a BCM-like rule. Compare to Fig.~\ref{fig:corr_dimension_scaling}.}
			\label{fig:corr_dimension_scaling_bcm}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

		\subsection{Objective Function of BCM Learning in the Compartment Model}
		\label{sect:Obj_Func}

		To form a better understanding of why the BCM-type learning rule
		in combination with the implemented compartment model, we can formalize
		the learning rule for the proximal weights in terms of an objective function.
		For this purpose, we further simplify \eqref{eq:comp_model} by replacing
		the sigmoid functions $\sigma(x)$ by a simple step function $\Theta(x)$. 
		This does not change the overall shape or topology of the activation
		in the $(I_p,I_d)$ space but merely makes the smooth transitions sharp
		and instantaneous. Using $\Delta w_i \propto y\left(y - \theta_M \right) x_i$,
		we find in this case
		\begin{equation}
			\begin{split}
			\Delta w_i \propto [ (1-\alpha) \Theta(I_d - \theta_{d})\Theta(p-\theta_{p1})
			\\ + \alpha (\alpha - 1)\Theta(\theta_{d} - I_d)\Theta(p-\theta_{p0}) ]x_i \; .
			\end{split}
		\end{equation}
		Noting that $\Theta(x)$ is the first derivative of the ReLu function $[x]^+ \equiv \max(0,x)$,
		we find that this update rule can be written as
		\begin{align}
			\Delta w_i &\propto \frac{\partial \mathcal{L}_p}{\partial w_i}\\
			\begin{split}
			\mathcal{L}_p &\equiv (1-\alpha) \Theta(I_d - \theta_{d})[p-\theta_{p1}]^+\\
			&\;\; + \alpha (\alpha - 1)\Theta(\theta_{d} - I_d)[p-\theta_{p0}]^+ \; .
			\end{split}\label{eq:obj_func}
		\end{align} 
		The objective function is shown in Fig.~\ref{fig:obj_func}. One can observe
		that states closer to the $I_p$-$I_d$ diagonal are preferred, while the opposite is the 
		case for off-diagonal states. This provides an explanation why the BCM-rule can induce 
		an alignment between proximal and distal inputs when used in combination with the nonlinear 
		compartment model. 
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\begin{figure}
			\includegraphics[width=\columnwidth]{obj_func}
			\caption{{\bf Objective Function for the Proximal Weight Update.} The 
			approximate objective function for the proximal weights as given in
			\eqref{eq:obj_func}. This corresponds to a combination of using 
		\eqref{eq:comp_model} together with \eqref{eq:bcm_rule}. Note the ridge-like
		structure along the $I_p$-$I_d$ diagonal, which supports the alignment between
		proximal and distal input.}
		\label{fig:obj_func}
		\end{figure}
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

		\section{Discussion}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.   
%                                              %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.4 Generated 2018/06/15 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%
%%%  NB logo1.jpg is required in the path in order to correctly compile front page header %%%

\documentclass[utf8]{frontiersSCNS} % for Science, Engineering and Humanities and Social Sciences articles
%\documentclass[utf8]{frontiersHLTH} % for Health articles
%\documentclass[utf8]{frontiersFPHY} % for Physics and Applied Mathematics and Statistics articles

%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}
\linenumbers

\graphicspath{{./figures/}}

\def\keyFont{\fontsize{8}{11}\helveticabold }

\def\firstAuthorLast{Fabian Schubert and Claudius Gros} %use et al 
%only if is more than 1 author

\def\Authors{Fabian Schubert\,$^{1,*}$ and Claudius Gros\,$^{1}$}

% Affiliations should be keyed to the author's name with superscript 
%numbers and be listed as follows: Laboratory, Institute, Department, 
%Organization, City, State abbreviation (USA, Canada, Australia), and 
%Country (without detailed address information such as city zip codes 
%or street names).
% If one of the authors has a change of address, list the new address 
%below the correspondence details using a superscript symbol and use 
%the same symbol to indicate the author in the author list.

\def\Address{$^{1}$Institute for Theoretical Physics, 
	Goethe University Frankfurt am Main, Germany}

% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name 
%and city zip code) and email of the corresponding author

\def\corrAuthor{Institute for Theoretical Physics\\
	Goethe University Frankfurt am Main\\
	Max-von-Laue-Str. 1\\
	60438 Frankfurt am Main, Germany}

\def\corrEmail{fschubert@itp.uni-frankfurt.de}

\begin{document}
\onecolumn
\firstpage{1}

\title[Running Title]{Nonlinear Dendritic Coincidence Detection for Supervised Learning}

\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

\maketitle

\begin{abstract}

Cortical pyramidal neurons have a complex dendritic anatomy, whose
function is an active field of scientific research.
In particular, the segregation between its soma and the apical dendritic tree
is believed to play an active role in processing feed-forward 
sensory information and top-down or feedback signals. In this work, we use a simple 
two-compartment model accounting for the nonlinear interactions between
basal and apical input streams and show that a simple Hebbian learning rule
in the basal compartment allows the neuron to align its basal input to 
a target signal in the apical compartment. We show that this learning
process is robust against strong distractions in the basal input space
and demonstrate its effectiveness in a linear classification task.

\tiny
 \keyFont{ \section{Keywords:} Dendrites, Pyramidal Neuron, Plasticity, 
 	Coincidence Detection, Supervised Learning}
\end{abstract}

%---------------------------------------%
\section{Introduction}
\label{sect:introduction}
%---------------------------------------%

@@

%---------------------------------------%
\section{Model}
\label{sect:model}
%---------------------------------------%

%---------------------------------------%
\subsection{Neuron Model}
\label{sect:neuronmodel}
%---------------------------------------%

The neuron model used throughout this study is a discrete-time 
rate encoding model that uses two separate input variables, 
subsuming the total synaptic input current injected arriving 
at the basal (proximal) and apical (distal) 
dendritic structure of a pyramidal neuron, respectively. 
The model is a slightly simplified version of a phenomenological 
model proposed by \citet{Shai_2015}. Denoting the input currents 
$I_{\rm p}$ (proximal) and $I_{\rm d}$ (distal), 
the model is written as
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
\begin{split}
y\left(t\right) &= \alpha  \sigma\left( I_{\rm p}(t) - \theta_{p0} \right)
\left[1-\sigma\left(I_{\rm d}(t) - \theta_d\right)\right] \\
&+ \sigma\left(I_{\rm d}(t) - \theta_d \right)
\sigma\left( I_{\rm p}(t) - \theta_{p1} \right)
\end{split} \label{eq:comp_model}\\
\sigma(x) &\equiv \frac{1}{1+\exp(-4x)} \; .
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here, $\theta_{p0}$, $\theta_{p1}$ and $\theta_d$ are threshold 
variables with respect to proximal and distal input. 
Overall, this equation describes two distinct
regions of neural activation in the 
$(I_{\rm p}, I_{\rm d})$-space which differ in their
maximal firing rates, which are set to $1$ and 
$\alpha$, where $0 < \alpha < 1$.
A plot of \eqref{eq:comp_model} is shown in Fig.~\ref{fig:comp_model}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.6\columnwidth]{plot_comp_mod_marks.png}
\caption{{\bf Two-compartment rate model.} Firing rate as a
	function of proximal and distal input $I_{\rm p}$ and $I_{\rm d}$, 
	see \eqref{eq:comp_model}. 
	The thresholds $\theta_{p0}$, $\theta_{p1}$ and $\theta_d$ define
	two regions of neural activity with a a maximal firing rate of $1$
	and $\alpha$.}
\label{fig:comp_model}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%---------------------------------------%
\subsection{Plasticity}
%---------------------------------------%

We implemented a Hebbian plasticity rule for the basal 
synaptic weights given by the following update equation:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
w_i(t+1) &= w_i(t) + \mu_w \left(x_{p,i}(t) - \tilde{x}_{p,i}(t)\right)
\left(y(t) - \tilde{y} \right)
\label{eq:hebb_plast} \\
\tilde{x}_{p,i}(t+1) &= (1-\mu_{av})\tilde{x}_{p,i}(t) + \mu_{av}x_{p,i}(t+1) \\
\tilde{y}(t+1) &= (1-\mu_{av})\tilde{y}(t) + \mu_{av}y(t+1)
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Additionally, we use a
synaptic normalization constraint
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
w_i(t) \rightarrow \frac{w_i(t)}{\lVert \mathbf{w}(t)\rVert}
\label{eq:weight_norm}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
in each time step, where $\lVert \mathbf{w}(t)\rVert$ denotes the Euclidean
norm of the synaptic weight vector.

For comparative reasons, the point neuron model is equipped with the
same plasticity rule for the proximal weights as \eqref{eq:hebb_plast}.

@@

%---------------------------------------%
\section{Results}
\label{sect:results}
%---------------------------------------%

%---------------------------------------%
\subsection{Increased Alignment between Basal and Apical Inputs}
\label{sect:alignment}
%---------------------------------------%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.7\columnwidth]{corr_dimension_scaling_high_input_dim}
\caption{{\bf Alignment between Basal and Apical Input.} Color
	encodes the Pearson correlation $\rho[I_p,I_d]$ for different
	number of orthogonal distraction directions $N_{dist}$ 
	and the corresponding scaling faction $s$.}
\label{fig:corr_dimension_scaling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%---------------------------------------%
\subsection{Supervised Learning in a Linear Classification Task}
\label{sect:classification}
%---------------------------------------%

Next, we investigated if the observed differences would also improve
the performance in an actual supervised learning task.
For this purpose, we constructed presynaptic basal 
input as illustrated in Fig.~\ref{fig:illustration_classification}.
Written in vector form, each sample from the basal 
input is generated from the following expression:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\mathbf{x}_p(t) = \mathbf{b} + \mathbf{a}\left(c(t) + \sigma_a \zeta_a(t) \right) 
+ s \cdot \sum_{i=1}^{N_{dist}} \zeta_{dist,i}(t) \mathbf{v}_{dist,i} \; .
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here, $\mathbf{b}$ is a random vector drawn uniformly from
$(0,1)^N$, $\mathbf{a}$ is random unit vector as introduced in 
Section~\ref{sect:alignment}, $c(t)$ is a binary variable drawn 
from $\{-0.5,0.5\}$ with equal probability and $\zeta_a(t)$ and the
$\zeta_{dist,i}(t)$ are independent random Gaussian variables with 
zero mean and unit variance. 
Hence, $\sigma_a$ simply denotes the standard deviation of each Gaussian
cluster along the direction of the normal vector $\mathbf{a}$ and was
set to $\sigma_a = 0.25$. 
Finally, the set of $\mathbf{v}_{dist,i}$ forms a randomly generated
orthogonal basis of $N_{dist}$ unit vectors which---just as in 
Section~\ref{sect:alignment}---are also orthogonal to $\mathbf{a}$.
The free parameter $s$ parameterized
the standard deviation along this subspace orthogonal to $\mathbf{a}$.
As indicated by the time dependence, the Gaussian and binary random
variables are drawn for each time step. The vectors
$\mathbf{b}$, $\mathbf{a}$, and $\mathbf{v}_{dist,i}$ are generated
once before the beginning of a simulation run.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.55\columnwidth]{illustration_classification}
\caption{{\bf Input Space for a Simple Linear Classification Task.}
	Two clusters of presynaptic basal activity were generated from 
	multivariate Gaussian distributions. Here, $s$ denotes the standard
	deviation orthogonal to the normal vector of the classification
	hyperplane.}
\label{fig:illustration_classification}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the classification task, we set up two output neurons receiving
the same basal presynaptic input, while the top-down input encoded
the correct linear classification in a one-hot scheme, that is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
x_{d,0}(t) &= 1 - \Theta\left( \left(\mathbf{x}_p(t) - 
\mathbf{b}\right)^T \mathbf{a}\right) \\
x_{d,1}(t) &= \Theta\left( \left(\mathbf{x}_p(t) - 
\mathbf{b}\right)^T \mathbf{a}\right) \; ,
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\Theta(x)$ is the Heaviside step function.

As in the previous experiment, we ran a full simulation until 
all dynamic variables reached a stationary state. After this,
a test run without plasticity and with the apical input turned off 
was used to evaluate the classification	performance. 
For each sample, the index of the neuron with the highest 
activity was used as the predicted class. 
Accuracy was then calculated as the fraction
of correctly classified samples.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.7\columnwidth]{classification_dimension_scaling_high_input_dim}
\caption{{\bf Binary Classification Accuracy.}
	Fraction of correctly classified patterns as illustrated in
	Fig.~\ref{fig:illustration_classification}, see 
	Section~\ref{sect:classification}.¸}
\label{fig:classification_accuracy}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The resulting accuracy as a function of $N_{dist}$ and $s$
is shown in Fig.~\ref{fig:classification_accuracy}.
Albeit differences are small, the compartment model does
show a better overall accuracy for the tested parameter range.
It should be noted, though, that the advantage of the compartment
model becomes even less prominent when looking at the actual
correlation between proximal and distal input as a
measure of successful learning (as done in the previous section).
Fig.~\ref{fig:classification_correlation} in the supplementary 
section shows that the compartment model allows for a marginally
larger scaling among the distraction dimensions.
\\
%---------------------------------------%
\subsection{Non-Hebbian Learning Rules}
\label{sect:non-hebbian}
%---------------------------------------%

@@

The compartment model also significantly benefits from
the BCM rule in terms of basal-apical alignment as tested in 
Sect.~\ref{sect:alignment}, while only marginal
improvements can be observed for the point model
(compare Fig.~\ref{fig:corr_dimension_scaling_bcm} 
with Fig.~\ref{fig:corr_dimension_scaling}).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.7\columnwidth]{classification_dimension_scaling_bcm_high_input_dim}
\caption{{\bf Binary Classification Accuracy, BCM Rule.}
	Fraction of correctly classified patterns as illustrated in
	Fig.~\ref{fig:illustration_classification}, after training with
	a BCM-like learning rule.¸}
\label{fig:classification_accuracy_bcm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.7\columnwidth]{corr_dimension_scaling_bcm_high_input_dim}
\caption{{\bf Alignment between Basal and Apical Input, BCM Rule.}
	Color encodes the Pearson correlation $\rho[I_p,I_d]$ for different
	number of orthogonal distraction directions $N_{dist}$ 
	and the corresponding scaling faction $s$ after training
	with a BCM-like rule. Compare to Fig.~\ref{fig:corr_dimension_scaling}.}
\label{fig:corr_dimension_scaling_bcm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\
\\
%---------------------------------------%
\subsection{Objective Function of BCM Learning in the Compartment Model}
\label{sect:Obj_Func}
%---------------------------------------%

To form a better understanding of why the BCM-type learning rule
in combination with the implemented compartment model, we can formalize
the learning rule for the proximal weights in terms of an objective function.
For this purpose, we further simplify \eqref{eq:comp_model} by replacing
the sigmoid functions $\sigma(x)$ by a simple step function $\Theta(x)$. 
This does not change the overall shape or topology of the activation
in the $(I_p,I_d)$ space but merely makes the smooth transitions sharp
and instantaneous. Using $\Delta w_i \propto y\left(y - \theta_M \right) x_i$,
we find in this case
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{split}
\Delta w_i \propto [ (1-\alpha) \Theta(I_d - \theta_{d})\Theta(p-\theta_{p1})
\\ + \alpha (\alpha - 1)\Theta(\theta_{d} - I_d)\Theta(p-\theta_{p0}) ]x_i \; .
\end{split}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Noting that $\Theta(x)$ is the first derivative of the ReLu function $[x]^+ \equiv \max(0,x)$,
we find that this update rule can be written as
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{align}
\Delta w_i &\propto \frac{\partial \mathcal{L}_p}{\partial w_i}\\
\begin{split}
\mathcal{L}_p &\equiv (1-\alpha) \Theta(I_d - \theta_{d})[p-\theta_{p1}]^+\\
&\;\; + \alpha (\alpha - 1)\Theta(\theta_{d} - I_d)[p-\theta_{p0}]^+ \; .
\end{split}\label{eq:obj_func}
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The objective function is shown in Fig.~\ref{fig:obj_func}. 
One can observe that states closer to the $I_p$-$I_d$ diagonal 
are preferred, while the opposite is the case for off-diagonal states. 
This provides an explanation why the BCM-rule can induce 
an alignment between proximal and distal inputs when used 
in combination with the nonlinear compartment model. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.7\columnwidth]{obj_func}
\caption{{\bf Objective Function for the Proximal Weight Update.} The 
	approximate objective function for the proximal weights as given in
	\eqref{eq:obj_func}. This corresponds to a combination of using 
	\eqref{eq:comp_model} together with \eqref{eq:bcm_rule}. Note the ridge-like
	structure along the $I_p$-$I_d$ diagonal, which supports the alignment between
	proximal and distal input.}
\label{fig:obj_func}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%---------------------------------------%
\section{Discussion}
%---------------------------------------%
